%\index{\footnote{}}
\chapter{Introduction}
Developing a unified architectural pipeline for multi-class abnormality classification and grounded localization
in 3D Computed Tomography (CT)a ddresses two of the most critical imperatives in clinical AI adoption: diagnostic accuracy and explainability.
Specifically, utilizing dense pixel-level segmentation as the mechanism for localization forces the architecture to
provide voxel-level visual evidence for its diagnostic claims. This ensures that the model is held accountable and its
reasoning aligns with clinical diagnosis.

\section{Background}
The transition from 2D projectional radiography to volumetric CT represents a paradigm shift in medical image analysis.
For Lung disease diagnosis, Chest X-rays (CXR) are commonly used as the first imaging test due to its low cost and low
radiation exposure. However, studies show that their capability to detect abnormailities is limited when compared to 3D CT.
\cite{SELF2013401}.

CXR has historically benefited from the availability of massive, radiologist-annotated datasets, while
the analysis of 3D chest CT volumes has been severely constrained by a lack of such data.
However, the recent introduction of large-scale, text-paired datasets has facilitated the development of state-of-the-art
(SotA) vision-language foundation models tailored specifically for volumetric imaging.
\cite{baharoon2025rexgroundingct3dchestct} \cite{doi:10.1200/JCO.2016.34.15\_suppl.e18246} \cite{chen2026dflashblockdiffusionflash}.

Beyond deep learning approaches, the Label Consistent K-SVD (LC-KSVD)
algorithm offers a viable approach for interpretable CT analysis. LC-KSVD is a dictionary learning algorithm that
incorporates label information directly into the dictionary construction process \cite{LCKSVD}, making it inherently discriminative and
well-suited for multi-abnormality tasks.

\section{Problem Statement}
The clinical utility of Deep Learning-based approaches is hindered by a lack of interpretability. Tools like Grad-CAM
provide visual heatmaps highlighting important regions in an input image by analyzing gradients from the final convolutional
layer \cite{Suara_2024}. These post-hoc explanations are often insufficient for medical professionals who require a granular understanding
of the model's decision making process. In high stakes environments, a model that produces a diagnosis without transparency
in its reasoning is difficult to integrate into standard workflows, as it lacks the accountability necessary for legal
and ethical compliance.

\section{Objectives and Scope}
The primary goal of this project is to address the limitations of existing Deep Learning based approaches for Lung abnormality
classification and localization by developing an inheerntly interpretable dictionary learning framework based on LC-KSVD.
This approach aims to enable transparent, anatomically grounded decision making, where each classification outcome
can be traced back to specific pathological patterns. This aligns with clinical reasoning and facilitates validation and
adoption in radiological workflows. The specific objectives are as follows:

\subsection{Objectives}
\begin{itemize}
	\item Implement a Dictionary Learning approach to learn normal and pathological dictionary atoms using a publicly available annotated 3D chest CT dataset.
	\item Map dictionary atoms back to anatomical features and visualize their spatial contribution maps.
	\item Benchmark model performance against state-of-the-art classification and segmentation models on the same dataset.
	\item Evaluate abnormality localization by the model on a real-world dataset with expert radiologist review.
\end{itemize}

\subsection{Scope}
The scope of this study focuses on the development and evaluation of a dictionary learning based approach for classification
and grounded localization of lung abnormalities using volumetric 3D CT data, with the following limitations:

\begin{itemize}
	\item This study is limited to 3 thoracic abnormalities: pulmonary nodules, ground-glass opacities, and consolidations.
	\item 
	\item 
\end{itemize}

\section{Literature Review}
The landscape of 3D CT analysis is currently divided into three primary architectural paradigms:
\begin{itemize}
	\item Classification-based models that prioritize diagnostic labels
	\item Segmentation-based models that prioritize boundary precision
	\item Unified architectures designed for grounded localization
\end{itemize}.

SotA models are predominantly specialized in either classification or segmentation. However, recent systems, particularly
in 2025 and 2026, are moving toward multi-task learning or foundation models that can perform both diagnosis and voxel-level
segmentation simultaneously.

The efficacy of these models is increasingly evaluated on large-scale, multi-institutional benchmarks such as CT-RATE,
FLARE 2024-2025 and MSWAL, which provide the complexity necessary to test generalizability across diverse patient populations
and scanner manufacturers.

\subsection{Classification Architectures}
Classification-based models for 3D CT have shifted away from 3D Convolutional Neural Networks (CNNs) such as 3D U-Net,
3D ResNet and DenseNet pipelines, toward Vision Transformers (ViTs) and self-supervised encoders. These models are designed
to identify the presence of one or more abnormalities
within a full volumetric scan. The benchmark for this paradigm is the CT-RATE dataset, which introduced the first public
large-scale 3D medical imaging dataset with paired textual reports, enabling the development of Contrastive Language-Image
Pretraining for Computed Tomography (CT-CLIP).

(About CT-CLIP)
Recent benchmarks such as Merlin and DINOv3 have further challenged the necessity of domain-specific pretraining.

Merlin is a vision-language foundation model exclusively pre-trained on abdominal CT scans. Despite it's training, Merlin
has consistently demonstrated exceptional zero-shot performance on chest CT classification tasks, even when it was
restricted to a linear probe, significantly outperforming models like CT-CLIP, M3FM, and CT-FM that were explicitly
trained on thoracic data, by an average of 12.3\% in Area Under the Receiver Operating Characteristic Curve (AUROC).
Specifically, in cross-domain retrieval tasks, Merlin surpassed CT-CLIP by 24.7\%.
DINOv3, a self-supervised ViT pretrained on over 1.7 billion natural images, outperform specialized medical models like
CT-Net and CT-CLIP across all metrics on several classification tasks due to it's ability to capture fine-grained textural features.

\subsection{Segmentation Architectures}
The prevailing SotA for supervised volumetric segmentation remains variations of the 3D U-Net, such as nnU-Net, which
utilizes an adaptive framework to optimize preprocessing and architecture for specific datasets. However, the field has
recently shifted toward foundation models capable of universal segmentation such as SegVol and SAM-Med3D.
The SegVol model is the first 3D foundation model specifically designed for universal volumetric medical image segmentation.
Similarly, SAM-Med3D adapts the Segment Anything Model (SAM) for 3D medical data.

Recent evaluations published at the Machine Learning for Healthcare (MLHC) Conference in 2025, which specifically utilized
the ReXGroundingCT dataset, revealed that the current SotA text-prompted segmentation models that demonstrate near-perfect
performance in general computer vision tasks struggle with the nuanced clinical descriptions and ill-defined boundaries
of chest CT abnormalities.

\subsection{Unified Architectures}
Recent models have been designed to integrate abnormality detection, segmentation, and multimodal chain-of-thought reasoning,
enabling pixel-level localization, structured report generation, and physician-like diagnostic inference in a single framework.
BiomedParse, and it's production-ready, version MedImageParse developed by Microsoft, are foundation models designed to parse 3D
medical images.
Other notable models include Citrus-V, MTMed3D and MedVista3D, where each architecture incorporates a multi-task learning
(MTL) framework.

\subsection{Dictionary Learning Approaches}
Applications of sparse dictionary learning in medical imaging have demonstrated significant success across diverse diagnostic tasks. In histopathological
analysis, Discriminative Feature oriented Dictionary Learning (DFDL) has shown the ability to learn class specific dictionaries that achieve high classification accu-
racy even with limited training data a critical advantage when generous training samples are not available[19]. In neuroimaging, multi feature kernel supervised
dictionary learning approaches have demonstrated superior performance in distinguishing Alzheimer’s disease patients from cognitively unimpaired individuals,
achieving 98.18\% classification accuracy while maintaining fast testing times[20].

Most relevant to our work, dictionary learning has been successfully applied to lung disease diagnosis from CT images. One approach utilized both frozen dic-
tionary learning and label consistent LC-KSVD to classify COVID 19 CT scans, achieving 89\% accuracy on the CC-CCII dataset[21]. The framework’s key ad-
vantage lies in its interpretability, where the sparse coding output can be directly mapped back to the original input signal domain. This work demonstrates that
sparse representation methods offer an effective balance between performance and explainability for lung disease classification tasks, providing clear feature represen-
tations while maintaining competitive accuracy. Similarly, sparse representation methods have achieved 95.4\% accuracy in classifying diffuse lung disease pat-
terns on HRCT images, effectively handling the high variety and complexity of patterns[22].


\subsection{Explainability Approaches}

Explainability methods are essential in chest CT analysis because abnormality classification alone is not sufficient for clinical adoption. 
For this project, the most relevant explainability techniques are those that can relate predictions back to anatomical regions and, ideally, support voxel-level interpretation in volumetric data.

Grad-CAM is one of the most widely used post-hoc explanation methods for CNN-based classifiers. 
It generates class-discriminative heatmaps by using gradients flowing into the final convolutional layer, allowing the user to identify image regions that influenced a prediction. 
In medical imaging, Grad-CAM has been applied to chest radiographs and CT classification tasks to provide coarse visual explanations. 
However, its main limitation is that it produces low-resolution saliency maps and does not directly provide precise lesion boundaries or voxel-level localization, which is a major drawback for 3D CT interpretation.

Seg-Grad-CAM extends this idea by incorporating segmentation information into the explanation process. 
Instead of only highlighting regions that are important for classification, it attempts to align attribution maps with anatomical or lesion masks, making the explanation more clinically meaningful. 
This is particularly useful for volumetric medical imaging tasks where the target abnormality occupies a sparse region within a large 3D scan. 
For a project focused on grounded localization, Seg-Grad-CAM is more relevant than standard Grad-CAM because it better reflects whether the model is attending to the pathological structure rather than surrounding tissue.

HiRes-CAM addresses one of the key weaknesses of Grad-CAM by producing higher-resolution localization maps. 
This is important in chest CT because many abnormalities, such as small nodules, consolidations, or subtle interstitial changes, may be missed by coarse heatmaps. 
HiRes-CAM aims to preserve fine spatial detail, making it more suitable for medical imaging applications where precise localization is required. 
Nevertheless, like other gradient-based methods, it remains a post-hoc explanation technique and does not change the internal decision process of the model.

In the context of this project, these methods are useful baselines for evaluating interpretability in deep learning models. 
However, they remain fundamentally different from dictionary learning approaches such as LC-KSVD, which provide intrinsic interpretability by representing predictions through sparse combinations of learned atoms. 
Compared with Grad-CAM, Seg-Grad-CAM, and HiRes-CAM, LC-KSVD offers a more direct and anatomically traceable explanation mechanism for multi-abnormality classification and localization in 3D CT.

\section{Report Outline}



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%% END OF FILE
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
