\chapter{Methodology}

\section{Data Acquisition and Preprocessing}
Two public chest CT datasets were used in this work: CT-RATE and ReXGroundingCT. 
CT-RATE was used as the primary source of volume-level labels (normal/abnormal and abnormality presence) derived from the dataset provided multi finding annotations, 
while ReXGroundingCT was used to obtain lesion level spatial supervision through radiologist linked segmentation masks together with finding descriptions. 
In practice, CT-RATE provided the global abnormality labels needed to build classification splits, and ReXGroundingCT provided grounded masks and metadata required for localisation-oriented experiments.

After an initial review of the 18 thoracic findings available in CT-RATE, a clinical discussion with the supervising radiologist produced a shortlist of abnormalities considered both clinically meaningful and suitable for interpretable analysis 
(Atelectasis, Lung nodule, Ground-Glass Opacity (GGO), Consolidation, Mosaic attenuation, Bronchiectasis, and Calcification). 
However, after checking practical label and mask availability across the two datasets, the final experiments were restricted to three abnormalities with the most consistent coverage: \textit{Lung nodule}, \textit{Lung opacity / GGO}, and \textit{Consolidation}. 
The remaining candidates were excluded mainly due to limited grounded lesion samples, inconsistent annotation coverage, or insufficiently distinct mask supervision for reliable training and evaluation.

The preprocessing pipeline followed the LC-KSVD implementation. 
CT volumes and corresponding masks were loaded from NIfTI files, then resampled to an isotropic spacing of 1.5\,mm to reduce inter-scan resolution variability. 
Volumes were resampled with trilinear interpolation and segmentation masks with nearest-neighbour interpolation to preserve discrete labels. 
Intensities were windowed using a lung HU range of $[-1000, 200]$ and linearly normalised to $[0,1]$. 
For dictionary learning, each preprocessed volume was converted into fixed-size 3D patches of $32\times32\times32$ voxel. 
Positive patches were sampled from lesion regions using the grounded masks (minimum overlap ratio 5\%, with a fallback for very small lesions). 
Negative patches were sampled from non overlapping regions (including normal scans), and patch vectors were $L_2$-normalised prior to sparse coding.


\section{Implementation of the LC-KSVD Approach}
To implement the Dictionary Learning approach for our study, we required a concrete implementation of the LC-KSVD algorithm.
Currently, many of the available standard implementations of KSVD and LC-KSVD are developed using MATLAB, which is not ideal
for integration with our Python-based data processing and evaluation pipeline. Although A few open-source Python implementations
LC-KSVD are available, they either lacked an importable library, or did not provide the flexibility required for our study.

Therefore, we developed a custom implementation of the LC-KSVD algorithm in Python and published it on PyPI under the name “reppi.”
The implementation follows a modular design, consisting of three separate components:
\begin{enumerate}
    \item SRC - Utility Functions shared across Sparse Coding Algorithms
    \item OMP - Implementation of Batch-OMP and OMP-Cholesky approaches for Sparse Coding
    \item KSVD - Implementation of KSVD Dictionary Learning Algorithm
    \item LC-KSVD - Implementation of LC-KSVD1 and LC-KSVD2 Dictionary Learning Algorithms
\end{enumerate}

This design allows for flexibility in integrating alternative sparse coding methods. The source code is available at github.com/ckekula/reppi.

This modular architecture enables straightforward integration of the algorithm through library imports, while providing
the flexibility required to experiment with different hyperparameters and configurations. The implementations of the SRC,
OMP, and LC-KSVD modules are provided in Appendix A, Appendix B, and Appendix C, respectively.

\section{Abnormality Classification}
Abnormality classification was implemented as a set of one-vs-normal binary LC-KSVD2 models, trained separately for each target finding (Lung nodule, Lung opacity, and Consolidation). 
After preprocessing, each CT volume is represented as a matrix of vectorised $32\times32\times32$ patches, with each patch column-wise normalised to unit $L_2$ norm before sparse coding. 
At inference time, patch-wise sparse codes $\Gamma$ are computed using OMP with the learned dictionary $D$, and classification is performed using the learned linear classifier weights $W$ from LC-KSVD2 by scoring $s = W_{\text{pos}}\,\Gamma$. 
For volume-level prediction, the implementation aggregates evidence by averaging the sparse codes across all sampled patches and applies a sign threshold ($s>0$) to obtain the final abnormality present/absent decision.

\section{Back-Projecting Class-Discriminative Atoms}
The unique intuition we propose in our approach is to generate spatial contribution maps through back-projection. Because
each atom in the dictionary $D$ is associated with a specific class label through the matrix $Q$, the model can theoretically
"explain" its decision for a test patch by examining which atoms were used in its sparse representation.

To generate a spatial contribution map for a class $c$, the algorithm identifies all atoms $d_k$ that have been assigned
to that class. For a query voxel or patch with sparse coefficients $x$, the contribution $C_c$ is calculated as the
partial reconstruction:

$$C_c = \sum_{k \in \text{Class } c} d_k x_k$$

This process should theoretically back-project the class-specific features into the original voxel space, creating a map
that highlights the regions contributing to the detection of a specific abnormality. for class-discriminative atoms
operating on raw voxel-space inputs, the back-projection is an exact, closed-form spatial attribution.


\section{Deriving Explainability through Segmentation Masks}
To evaluate the clinical accuracy of LC-KSVD, these contribution maps are converted into binary segmentation masks for
comparison with ground truth annotations. This conversion can be achieved through various thresholding methods.

Grad-CAM's approximation has two sources of imprecision — the gradient itself is a local linear approximation of a non-linear
function, and the spatial upsampling from the coarse feature map to input resolution introduces further interpolation error.
LC-KSVD's contribution map has only one source of imprecision: the thresholding step you apply to binarize it into a mask.
The map itself — before thresholding — is exact given the sparse code and the dictionary.


Unlike the passive Grad-CAM approach, atom back-projection is generative. The atoms are what the model actually used to
reconstruct and classify the signal. There's no proxy involved. The localization evidence is produced by the same
computation that produced the classification, not by a separate post-hoc analysis run afterwards.


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%% END OF FILE
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
