\chapter*{Abstract}
Lung disease diagnosis from 3D chest CT scans faces critical challenges in computational efficiency and clinical interpretability.
Traditional 3D Convolutional Neural Networks require extensive labeled data and GPU resources, while post-hoc explainability methods
like Grad-CAM provide insufficient granularity for medical decision making.
While deep learning models achieve high accuracy, their black box nature and substantial computational demands limit clinical
adoption. This project proposes a sparse dictionary learning framework that addresses these limitations through inherently
interpretable feature representations. The methodology uses frozen dictionary learning approach combined with
Label-Consistent K-SVD (LC-KSVD) to capture both normal anatomical patterns and disease-specific pathological features. Dictionary
atoms are mapped to clinically meaningful structures such as Ground-Glass Opacity (GGO) and consolidation, enabling transparent
diagnostic reasoning. Each dictionary atom corresponds to specific volumetric patterns, allowing clinicians to trace classification
decisions back to anatomical features rather than relying on post-hoc visualization techniques.
The proposed system will be benchmarked against standard K-SVD and 3D CNN baselines using the CT-RATE and ReXGroundingCT datasets.
The sparse representation framework naturally handles class imbalance and requires fewer training samples than deep learning
approaches, making it particularly suited for rare pathologies. This research aims to demonstrate that sparse representation
methods can provide clinically interpretable lung disease classification while maintaining competitive performance with reduced computational overhead.
\newpage

% Table of Contents
\tableofcontents
\newpage

% List of Figures
\listoffigures
\newpage

% List of Tables
\listoftables
\newpage

% Acronyms
\addcontentsline{toc}{chapter}{Acronyms} % Add to table of contents
\acuseall % Use all acronyms to ensure they appear in the list
\printacronyms
\newpage