\chapter{Introduction}

\section{Background}
Lung disease remains a leading cause of global morbidity and mortality\cite{Zhai2025Global}. In the diagnosis of lung disease, assessing the pattern,
location, and regional distribution of involvement is the domain of radiology. Chest X-rays (CXR) are commonly used as the first 
imaging test due to its low cost and low radiation exposure. However, their diagnostic sensitivity is limited, as they capture
only a two-dimensional image of the chest. This causes anatomical superposition, where overlapping tissues and organs are projected
onto a single plane, allowing visualization of only about 70\% of the lung volume\cite{Wielpuetz2014}.

Computed Tomography (CT) is a radiographic technology that combines X-ray technology with computer reconstruction to produce 
images of cross sections of the human body\cite{Kasban2015}. This overcomes many of the limitations of radiography by eliminating
anatomical superposition and enabling full volumetric assessment of the lungs. CT produces high-resolution slices that reveal fine
structural details not visible in standard chest radiographs.

A defining characteristic of a CT scan is the Hounsfield Unit (HU), a quantitative scale of radiodensity that allows tissue properties
to be measured.
In pulmonary imaging, HU values allow for the objective differentiation of tissues such as normal parenchyma, Ground-Glass
Opacity (GGO), and consolidation that might appear visually similar. GGO appears as hazy increased opacity of lung, with
preservation of bronchial and vascular margins.
Consolidation is more opaque, and appears as a homogeneous increase in pulmonary parenchymal attenuation that obscures the margins
of vessels and airway walls\cite{Bankier2024}. While \ac{GGO} often suggests a reversible or early-stage condition, consolidation
usually indicates a more aggressive or advanced stage. The Consolidation-to-GGO (C/G) ratio is a quantitative metric calculated by dividing
the volume or area of consolidation by the volume or area of GGO within a lesion or throughout the lungs.

The traditional approach for CT analysis relies on a "slice-by-slice" approach where a radiologist views a series of axial
cross-sections and identifies the index slice (the single image where a pathology appears at its largest diameter)\cite{Gierada2016QuantitativeCT}.
A 2D slice only captures a small fraction of a 3D structure. If a nodule is irregularly shaped, the largest slice may not
represent the true extent of the disease. Studies show that 2D measurements are highly sensitive to the patient’s
orientation and the specific slice chosen, leading to high inter-observer variability. 2D models also treat each slice as an
independent data point, losing the "spatial continuity" between adjacent tissues.

3D volumetric analysis represents a quantitative shift wherein the entire dataset is used to reconstruct the lung's anatomy as a
solid volume. This enables more accurate characterization of lesion size, shape, and spatial distribution, while reducing
sensitivity to slice selection and viewing angle. 3D representations better capture disease morphology
and progression, providing a more robust foundation for quantitative analysis and automated learning-based methods.

Modern radiology increasingly relies on Artificial Intelligence (AI) to streamline the acquisition of radiologist analyses on
chest X-rays. However, the exponential increase in data volume generated by 3D CT scanners presents a profound challenge
for human interpretation and computational analysis. While the prevailing research trend has focused heavily on the deployment of
3D deep learning architectures, such as Convolutional Neural Networks (CNNs), these models frequently encounter significant hurdles
regarding computational overhead, data scarcity, and a critical lack of interpretability. The research proposed herein advocates
for a departure from these "black box" paradigms, suggesting that sparse representation methods offer a more intutive framework for the
volumetric identification of lung diseases by ensuring clinical interpretability.

\section{Problem Statement}
Modern radiology increasingly relies on AI to manage the vast quantities of 3D data produced during routine
screenings. 3D deep learning has been widely applied to the automatic diagnosis of lung diseases, achieving remarkable success in
tasks such as segmenting suspected infection regions and differentiating between various types of carcinoma. Despite these advancements,
the inherent complexity of 3D \ac{CNN}s introduces a set of constraints that are often overlooked in the pursuit of high accuracy scores.

\subsection{Computational Overhead}
Processing a 3D CT scan involves treating stacked 2D slices as a single volumetric matrix, which demands significantly more GPU
memory and processing power than standard image analysis. Unlike ordinary RGB images, CT scans consist of single-channel data with
varying voxel dimensions and resolution, which complicates spatial alignment and generalizability across different scanners.
Furthermore, the "curse of dimensionality" remains a persistent threat; as the network depth increases to capture global features,
the risk of gradient explosion and slow convergence becomes more pronounced.

\subsection{Interpretability Bottleneck}
Beyond computational limits, the clinical utility of deep learning is hampered by the interpretability paradox. While tools like
Grad-CAM provide visual heatmaps to localize lesions, these post-hoc explanations are often insufficient for medical professionals
who require a granular understanding of the model's decision-making process. In high-stakes environments, a model that shruggingly
produces a diagnosis without an audit trail of its reasoning is difficult to integrate into standard workflows, as it lacks the
accountability necessary for legal and ethical compliance. This transparency gap has led to a renewed interest in "open book" models
that are interpretable by design, where the internal logic is clear from the outset.

\section{Objectives and Scope}

\subsection{Objectives}

The primary goal of this project is to address the limitations of existing AI-based lung disease diagnosis systems, specifically
high computational overhead and lack of clinical interpretability, by developing an inherently interpretable and computationally
efficient dictionary learning framework. The proposed approach aims to enable transparent, anatomically grounded decision-making,
where each classification outcome can be traced back to specific pathological patterns. This aligns with clinical reasoning and
facillitates validation and adoption in radiological workflows.

\begin{enumerate}
    \item Develop an interpretable dictionary learning framework for lung disease classification from 3D chest CT scans
        \begin{itemize}
            \item Implement a two-phase training strategy to learn normal and pathological dictionary atoms
            using a publicly available annotated 3D chest CT dataset
            \item Integrate label-consistentency to improve class discrimination while maintaining
            interpretability
        \end{itemize}

    \item Improve model interpretability through visualization of learned dictionary atoms
        \begin{itemize}
            \item Visualize dictionary atoms in 3D and map them to anatomical and pathological lung features
            \item Analyze and compare sparse coefficient patterns across different disease classes
        \end{itemize}

    \item Evaluate the effectiveness of the proposed framework
        \begin{itemize}
            \item Benchmark performance against standard \ac{KSVD} without label consistency
            \item Compare results with a 3D \ac{CNN} baseline (3D-ResNet)
        \end{itemize}

    \item Integrate clinically relevant volumetric features into the analysis
        \begin{itemize}
            \item Perform HU-based segmentation to identify Ground-Glass Opacity (GGO) and consolidation
            \item Compute the Consolidation-to-GGO (C/G) ratio and relate model predictions to clinical thresholds
        \end{itemize}
\end{enumerate}


\subsection{Scope}
The scope of this study focuses on the development and evaluation of a dictionary learning-based framework
for lung disease classification using volumetric 3D chest CT data, with the following limitations:

\begin{itemize}
    \item The proposed solution is limited to volumetric CT scans of the thorax and does not consider other imaging modalities
    such as chest X-rays, MRI, or PET scans.
    \item The methodology is restricted to dictionary learning-based approaches, with particular emphasis on KSVD and LC-KSVD
    End-to-end deep dearning models such as CNNs are not the primary focus of this work and are included only as baseline
    comparators for performance benchmarking.
    \item Clinically, the analysis is confined to the characterization of lung parenchymal abnormalities, specifically GGO and
    consolidation patterns. Quantitative metrics such as HU-based segmentation and the C/G ratio are employed to support volumetric
    analysis. The study does not address disease prognosis or treatment outcome prediction.
    \item The scope excludes real-time clinical deployment, cross-institutional generalization studies, and regulatory validation.
    The findings are intended to demonstrate the feasibility and interpretability of sparse representation learning for 3D lung
    disease analysis rather than to provide a clinically deployable diagnostic system.
\end{itemize}
